---
output: github_document
always_allow_html: true
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# tines

<!-- badges: start -->
[![R-CMD-check](https://github.com/huizezhang-sherry/tines/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/huizezhang-sherry/tines/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/huizezhang-sherry/tines/graph/badge.svg)](https://app.codecov.io/gh/huizezhang-sherry/tines)
<!-- badges: end -->

The goal of tines is to ...

## Installation

You can install the released version of tines from CRAN with:

````
install.packages("tines")
````

And the development version from GitHub with:

````
# install.packages("remotes")
remotes::install_github("huizezhang-sherry/tines")
````

## Example - generate alternatives and code (HDI)

Start from a schema

```{r}
library(tines)
(hdi <- example_schema())
```

Generate an alternative at block-combine and write the result into a `yaml` file:

```{r eval = FALSE}
gen_alternatives(hdi, step = "step-combine", n = 1, 
                 file_path = here::here("inst/hdi-alt.yml"))
```


```{r}
#(res <- expand_tines(hdi, alternatives = here::here("inst/hdi-alt.yml")))
```

Take the original schema, original code, and the alternatives and generate a new R script that implements the alternative:

```{r eval = FALSE}
gen_code(hdi, 
         base_code = here::here("inst/hdi.R"), 
         alternative = here::here("inst/hdi-alt.yml"),
         output_dir = here::here("inst/")
         )
```

Run the generated code and compare the results:

```{r}
res_arith <- source(here::here("inst/alt_01_block_combine_arithmetic.R"))

res_arith$value |> 
  ggplot(aes(x = hdi, y = hdi2)) + 
  geom_point() + 
  theme(aspect.ratio = 1)
```


## Example - generate alternatives and code (football)

Currently restricted by a rate limit.

Start from a schema

```{r}
library(tines)
(football <- example_football())
```

Generate an alternative at block-combine and write the result into a `yaml` file:

```{r eval = FALSE}
gen_alternatives(football, id = "block-logistic-model", n = 2, 
                 file_path = here::here("inst/football-alt.yaml"))
```

Take the original schema, original code, and the alternatives and generate a new R script that implements the alternative:

```{r eval = FALSE}
gen_code(football, 
         base_code = here::here("inst/football.R"), 
         alternative = here::here("inst/football-alt.yaml"),
         output_dir = here::here("inst/")
         )
```

Run the generated code and compare the results:


```{r eval = FALSE}
schema <- example_rdi()
code <- gen_composite_code(
  schema       = schema,
   base_scripts = list(
     spei_template = here::here("inst/spei.R"),
     spi_template  = here::here("inst/spi.R")
  ),
  output_file  = here::here("inst/rdi.R")
 )

validate_script(here::here("inst/rdi.R"), verbose = TRUE)
```

## Example - generate alternatives and code (different random effect specifications)

```{r}
library(tidyverse)
library(broom.mixed)
```

### Generate a template to specify the alternatives

```{r eval = FALSE}
football_base <- read_tines(here::here("alternatives/football-grp5.yaml"))
# draft_alternatives(
#   x = football_base, 
#   id = "specify_random_effects_structure", 
#   output = here::here("alternatives/football-grp5-alt.yml")
# )
res <- expand_tines(football_base, here::here("alternatives/football-grp5-alt.yml"))
```

### Generate and validate the code for each alternatives
```{r eval = FALSE}
gen_code(res, output = here::here("alternatives/"), data = "data/football.csv")

files <- list.files(here::here("alternatives"), pattern = ".R", full.names = TRUE)[2:4]
map(files, ~validate_script(.x, data = here::here("data/football.csv")))
```

### Run the script for each alternatives and save the results for comparison
```{r eval = FALSE}
valid_files <- list.files(here::here("alternatives"), pattern = "iter2.R|football-grp5.R", full.names = TRUE)
mod_comp_res <- tibble(file = valid_files) |>  
  rowwise() |> 
  mutate(
    model = str_remove(basename(file), "_iter2.R|-grp5.R"),
    env = list(new.env()),
    res = list(source(file, local = env))
  )  
save(mod_comp_res, file = here::here("alternatives/mod_comp_res.rds"))
```

### Extract the model summary statistics and random effects for comparison
```{r eval = FALSE}
#load(here::here("alternatives/mod_comp_res.rds"))

map_dfr(mod_comp_res$env, ~glance(.x$glmm_model), .id = "model") |> 
  left_join(tibble(model = as.character(1:4), model2 = c("original", "gm1", "gm2", "gm3"))) |> 
  select(-c(model:logLik, df.residual)) |> 
  pivot_longer(cols = c(AIC, BIC, deviance), names_to = "measure", values_to = "value") |> 
  pivot_wider(names_from = model2, values_from = value)

map_dfr(mod_comp_res$env, ~tidy(.x$glmm_model, effect = "ran_pars", scales = "vcov"), .id = "model") |> 
  filter(!str_detect(term, "cov")) |> 
  left_join(tibble(model = as.character(1:4), model2 = c("original", "gm1", "gm2", "gm3"))) |> 
  select(-(model:component)) |> 
  pivot_wider(names_from = model2, values_from = estimate) |> 
  mutate(group = factor(group, levels = c("playerShort", "refNum", "refCountry"))) |> 
  arrange(group, term)

```
