---
output: github_document
always_allow_html: true
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# tines

<!-- badges: start -->
[![R-CMD-check](https://github.com/huizezhang-sherry/tines/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/huizezhang-sherry/tines/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/huizezhang-sherry/tines/graph/badge.svg)](https://app.codecov.io/gh/huizezhang-sherry/tines)
<!-- badges: end -->

The goal of tines is to ...

## Installation

You can install the released version of tines from CRAN with:

````
install.packages("tines")
````

And the development version from GitHub with:

````
# install.packages("remotes")
remotes::install_github("huizezhang-sherry/tines")
````

## Example - generate alternatives and code (HDI)

Start from a schema

```{r}
library(tines)
(hdi <- example_schema())
```

Generate an alternative at block-combine and write the result into a `yaml` file:

```{r eval = FALSE}
gen_alternatives(hdi, block = "block-combine", n = 1, 
                 file_path = here::here("inst/hdi-alt.yaml"))
```


```{r}
#(res <- expand_tines(hdi, alternatives = here::here("inst/hdi-alt.yaml")))
```

Take the original schema, original code, and the alternatives and generate a new R script that implements the alternative:

```{r eval = FALSE}
gen_code(hdi, 
         base_code = here::here("inst/hdi.R"), 
         alternative = here::here("inst/hdi-alt.yaml"),
         output_dir = here::here("inst/")
         )
```

Run the generated code and compare the results:

```{r}
res_arith <- source(here::here("inst/alt_01_block_combine_arithmetic.R"))

res_arith$value |> 
  ggplot(aes(x = hdi, y = hdi2)) + 
  geom_point() + 
  theme(aspect.ratio = 1)
```


## Example - generate alternatives and code (football)

Currently restricted by a rate limit.

Start from a schema

```{r}
library(tines)
(football <- example_football())
```

Generate an alternative at block-combine and write the result into a `yaml` file:

```{r eval = FALSE}
gen_alternatives(football, block = "block-logistic-model", n = 2, 
                 file_path = here::here("inst/football-alt.yaml"))
```

Take the original schema, original code, and the alternatives and generate a new R script that implements the alternative:

```{r eval = FALSE}
gen_code(football, 
         base_code = here::here("inst/football.R"), 
         alternative = here::here("inst/football-alt.yaml"),
         output_dir = here::here("inst/")
         )
```

Run the generated code and compare the results:

```{r}
```
