---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  message = FALSE,
  warning = FALSE
)
```

# tines <a href='https://huizezhang-sherry.github.io/tines/'><img src='man/figures/imgfile.svg' align="right" height="138.5" alt="tines package hex sticker logo featuring a fork-like branching diagram on a clean background, symbolising the multiverse of analytical paths the package helps explore" /></a>


<!-- badges: start -->
[![R-CMD-check](https://github.com/huizezhang-sherry/tines/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/huizezhang-sherry/tines/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/huizezhang-sherry/tines/graph/badge.svg)](https://app.codecov.io/gh/huizezhang-sherry/tines)
<!-- badges: end -->

The goal of tines is to help analysts document, explore, and communicate
the analytical decisions made during a data analysis. Analytical decisions
are recorded as a schema of steps, and large language models are used to
propose alternative choices at each step and to generate and validate the
corresponding R code. A schema can be expanded into a multiverse of
analyses, letting you see how your conclusions depend on the analytical
choices you made.


## Installation

You can install the released version of tines from CRAN with:

````
install.packages("tines")
````

And the development version from GitHub with:

````
# install.packages("remotes")
remotes::install_github("huizezhang-sherry/tines")
````

## LLM access

`extract_schema()`, `gen_alternatives()`, `gen_code()`, and `validate_script()`
all call out to an LLM via [ellmer](https://ellmer.tidyverse.org/). Set the
relevant API key as an environment variable before using them, e.g. for
Claude models:

````
usethis::edit_r_environ()
# add a line: ANTHROPIC_API_KEY=your-key-here
````

See `?ellmer::chat` for the full list of supported providers and their
expected environment variables.

## Example 1 - from a paper to validated code

### `extract_schema()` + `gen_code()` + `validate_script()`

`tines` can turn a methodology section from a paper into a schema, generate R
code that implements it, and validate that the code runs. The package ships
with the methodology text reported by Team 5 in the "many analysts, one
dataset" study of racial bias in football referee decisions
(Silberzahn et al., 2018), along with the football dataset they all analyzed.

```{r}
library(tines)
football_grp5
```

Extracting a schema from that text requires an LLM and a data dictionary:

```{r eval = FALSE}
data_dict <- c(
  "playerShort", "player", "club", "leagueCountry", "birthday", "height",
  "weight", "position", "games", "victories", "ties", "defeats",
  "goals", "yellowCards", "yellowReds", "redCards", "photoID", "rater1",
  "rater2", "refNum", "refCountry", "Alpha_3", "meanIAT", "nIAT",
  "seIAT", "meanExp", "nExp", "seExp"
)

extract_schema(football_grp5, data_dict,
               output_file = here::here("inst/football-grp5.yaml"))
```

```{r results = "asis", echo = FALSE}
cat("<details>\n<summary>You can click to expand the YAML file.</summary>\n\n")
cat("```yaml\n")
cat(readLines(here::here("inst/football-grp5/football-grp5.yaml")), sep = "\n")
cat("\n```\n\n</details>\n")
```

The YAML file can be loaded with `read_schema()`. Here the data is also stored in the package as 
`example_football_grp5()`:

```{r}
(schema <- example_football_grp5())
```

Generate an R script that implements the schema against the actual data:

```{r eval = FALSE}
gen_code(schema, data = here::here("inst/football.csv"),
         output = here::here("inst/football-grp5.R"))
```

```{r results = "asis", echo = FALSE}
cat("<details>\n<summary>You can click to expand the generated R script.</summary>\n\n")
cat("```r\n")
cat(readLines(here::here("inst/football-grp5/football-grp5.R")), sep = "\n")
cat("\n```\n\n</details>\n")
```

When the script generator didn't produce valid code in the first attempt, 
`validate_script()` can be used to iteratively fix it until it runs:

```{r eval = FALSE}
validate_script(here::here("inst/football-grp5.R"), data = here::here("inst/football.csv"))
```


## Example 2 - exploring alternatives

### `draft_alternatives()`/`gen_alternatives()` + `expand_tines()` + `gen_code()` + `validate_script()`

Given a schema, you may want to explore alternative analytical choices. 
You can write an alternative YAML file by hands from the template (`draft_alternatives()`) or 
ask LLM to propose alternatives for a step in the schema (`gen_alternatives()`).  

Here we have a pre-written alternative YAML file for the football example with different random effect specifications:

```{r results = "asis", echo = FALSE}
cat("<details>\n<summary>You can click to expand the alternative YAML file.</summary>\n\n")
cat("```yaml\n")
cat(readLines(here::here("inst/football-grp5/football-grp5-alt.yml")), sep = "\n")
cat("\n```\n\n</details>\n")
```

You can load an alternative YAML file with `read_alternatives()`:

```{r}
(alts <- read_alternatives(
  system.file("football-grp5/football-grp5-alt.yml", package = "tines")
))
```

Combine the alternatives with the original schema into a multiverse:

```{r}
(multiverse <- expand_tines(schema, alts))
```

After this, you can enerate code for every branch of the multiverse as we did for the original schema, and validate each generated script:

```{r eval = FALSE}
gen_code(multiverse, data = here::here("inst/football.csv"),
         output = here::here("inst/football-grp5-alternatives/"))

files <- list.files(here::here("inst/football-grp5-alternatives"),
                     pattern = "\\.R$", full.names = TRUE)
purrr::map(files, ~validate_script(.x, data = here::here("inst/football.csv")))
```


```{r results = "asis", echo = FALSE}
cat("<details>\n<summary>You can click to view the script generated for alternative gm1.</summary>\n\n")
cat("```yaml\n")
cat(readLines(here::here("inst/football-grp5/gm1_iter2.R")), sep = "\n")
cat("\n```\n\n</details>\n")
```

```{r results = "asis", echo = FALSE}
cat("<details>\n<summary>You can click to view the script generated for each alternative gm2.</summary>\n\n")
cat("```yaml\n")
cat(readLines(here::here("inst/football-grp5/gm2_iter2.R")), sep = "\n")
cat("\n```\n\n</details>\n")
```

```{r results = "asis", echo = FALSE}
cat("<details>\n<summary>You can click to view the script generated for each alternative gm3.</summary>\n\n")
cat("```yaml\n")
cat(readLines(here::here("inst/football-grp5/gm3_iter2.R")), sep = "\n")
cat("\n```\n\n</details>\n")
```
