---
title: "Structuring Alternative Files"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Structuring Alternative Files}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(tines)
```

An **alternatives file** describes one or more different ways to carry out
one or more steps in a `schema`. Passing one to `expand_tines()` turns a
single schema into a `multiverse`. This vignette covers

  * the anatomy of an alternatives YAML file,
  * how to create one by hand, in R, or with an LLM, and
  * how `expand_tines()` combines a schema and an alternatives file into a multiverse.

# The big picture

An alternatives file is a list of **nodes**. Each node names
a step (`overrides`) and a set of candidate `alternatives` for it, each with
its own `id`, `decision`, and `rationale` -- the skeleton always looks like
this:

```
meta:
  type: alternatives
  branch: single | multi
nodes:
  - overrides: <step id>
    alternatives:
      - id: <alternative id>
        decision: <...>
        rationale: <...>
      - id: <alternative id>
        ...
  - overrides: <step id>
    alternatives:
      ...
  ...
```

`meta.branch` tells `expand_tines()` how to turn the nodes into branches:

* **`single`**: every node must have exactly one alternative; they're all
  combined into a single branch that changes every listed step.
* **`multi`**: each node's alternatives are independent choices layered on
  top of original choice. Their cross-product is taken to produce a branch 
  for every combination of alternatives. For a node with `n` alternatives, 
  that's `n + 1` branches (the original plus one per alternative). For `K` 
  nodes with `n_i` alternatives each, that's `prod(n_i + 1)` branches.

# The HDI example

For the rest of this vignette, we'll use the Human Development Index (HDI) as an example. The HDI is a composite index of three dimensions: health, education, and standard of living that comprises of the following steps: 

  * `step-scaling`: Normalize the data to a common scale.
  * `step-education`: Calculate the education component of the HDI.
  * `step-combine`: Combine the three components into a single index.

```{r results = "asis", echo = FALSE}
schema_path <- system.file("hdi.yml", package = "tines")
cat("```yaml\n")
cat(readLines(schema_path), sep = "\n")
cat("\n```\n")
```

We can read this in and turn it into a `schema` object with `read_tines()`:

```{r}
schema <- read_tines(schema_path)
schema
```

# Case 1: Multiple steps on a single branch

Sometimes you want to create your own version of the analysis that mostly follows the original schema, but changes a few decisions -- for example, switching `step-scaling` to z-scoring and `step-education` to the max of the two school variables. `branch: single` combines those changes into one coordinated branch, rather than exploring each one independently:

```yaml
meta:
  type: alternatives
  branch: single
nodes:
  - overrides: step-scaling
    alternatives:
      - id: step-scaling-zscore
        decision: "z-score each variable"
        rationale: "puts variables on a common, interpretable scale"
  - overrides: step-education
    alternatives:
      - id: step-education-max
        decision: "use the max of the two school variables"
        rationale: "captures the higher-attainment measure"
```

In the above example, the `branch: single` indicates that both steps should be changed together in a single branch, rather than independently.

---

We can create this alternatives file in two ways: 
  
  * use `draft_alternatives()` to create a draft template and fill it in by hand, or
  * create it programmatically in R using the `node()`, `alternative()`, and `new_alternatives()` functions.

**1. Draft a template with `draft_alternatives()` and fill it in by hand:**

```{r}
draft_path_single <- withr::local_tempfile(fileext = ".yml")
draft_alternatives(
  schema,
  id = c("step-scaling", "step-education"),
  branch = "single",
  file_path = draft_path_single
)
```

```{r results = "asis", echo = FALSE}
cat("```yaml\n")
cat(readLines(draft_path_single), sep = "\n")
cat("\n```\n")
```

**2. Build one directly in R with `node()` + `alternative()` +
`new_alternatives()`:**

```{r}
alts_combo <- new_alternatives(
  node(
    overrides = "step-scaling",
    alternative(
      id = "step-scaling-zscore",
      decision = "z-score each variable",
      rationale = "puts variables on a common, interpretable scale"
    )
  ),
  node(
    overrides = "step-education",
    alternative(
      id = "step-education-max",
      decision = "use the max of the two school variables",
      rationale = "captures the higher-attainment measure"
    )
  ),
  branch = "single"
)
alts_combo
```

...and, if you want a YAML file out of it, write it back out with
`write_alternatives()`:

```{r}
tmp_file_single <- withr::local_tempfile(fileext = ".yml")
write_alternatives(alts_combo, tmp_file_single)
```

---

You can combine this alternative specification with the original schema to form a set of new schemas, or a **multiverse**, with 2 branches using `expand_tines()`:

```{r}
mv <- expand_tines(schema, alternatives = alts_combo)
mv
mv$`step-scaling-zscore+step-education-max`
```

Under `branch: single`, every node must have exactly one alternative.
This is checked as soon as an `alternatives` object is built, read, and printed:

```{r, error = TRUE}
alts_bad <- alts_combo
alts_bad$alternatives[[1]] <- rbind(
  alts_bad$alternatives[[1]],
  tibble::tibble(
    id = "step-scaling-minmax", decision = "...", rationale = "..."
  )
)
alts_bad
```

# Case 2: Checking multiple alternatives of a single step

One may consider testing a set of different alternatives for a single step. For example, one could use an arithmetic mean or a weighted mean that down-weights noisier dimensions. This can be represented in an alternatives file as follows:

```yaml
meta:
  type: alternatives
  branch: multi
nodes:
  - overrides: step-combine
    alternatives:
      - id: step-arithmetic-mean
        decision: "use an arithmetic mean"
        rationale: "the old method"
      - id: step-weighted-mean
        decision: "use a weighted mean, weighting each dimension by its variance"
        rationale: "down-weights noisier dimensions"
```

We can create this alternatives file in two ways:

* use `draft_alternatives()` to create a draft template and fill it in by hand, or
* create it programmatically in R using the `node()`, `alternative()`, and `new_alternatives()` functions.
* ask an LLM to propose alternatives with `gen_alternatives()`.

**1. Draft a template with `draft_alternatives()` and fill it in by hand:**

```{r}
draft_path <- withr::local_tempfile(fileext = ".yml")
draft_alternatives(
  schema, id = "step-combine", branch = "multi", file_path = draft_path
)
```

```{r results = "asis", echo = FALSE}
cat("```yaml\n")
cat(readLines(draft_path), sep = "\n")
cat("\n```\n")
```

`draft_alternatives()` requires a `branch` and `id` argument. The draft template 
contians two placeholder alternatives for one nodes. Users can edit the file to fill in the decision and rationale for each alternative, and add more alternatives or nodes as needed.

**2. Build one directly in R with `node()` + `alternative()` +
`new_alternatives()`:**

`new_alternatives()` likewise requires `branch` to be given explicitly:

```{r}
alts <- new_alternatives(
  branch = "multi",
  node(
    overrides = "step-combine",
    alternative(
      id = "step-arithmetic-mean",
      decision = "use an arithmetic mean",
      rationale = "the old method"
    ),
    alternative(
      id = "step-weighted-mean",
      decision = "use a weighted mean, weighting each dimension by its variance",
      rationale = "down-weights noisier dimensions"
    )
  )
)
alts
```

...and, if you want a YAML file out of it, write it back out with
`write_alternatives()`:

```{r}
tmp_file <- withr::local_tempfile(fileext = ".yml")
write_alternatives(alts, tmp_file)
```

```{r results = "asis", echo = FALSE}
cat("```yaml\n")
cat(readLines(tmp_file), sep = "\n")
cat("\n```\n")
```

**3. Ask an LLM to propose alternatives with `gen_alternatives()`:**

```{r, eval = FALSE}
gen_alternatives(
  schema,
  step = "step-combine", n = 3,
  file_path = "step-combine-alt.yml"
)
```

`gen_alternatives()` always targets one step, so it produces a `branch:
multi` file with exactly one node.

--- 

When `branch = multi`, `expand_tines()` produces a branch for every alternative, plus the original. In this case, with 2 alternatives, this results in 3 branches: 
```{r}
mv <- expand_tines(schema, alts)
mv
mv$`step-weighted-mean`
```

`expand_tines()` also accepts a path to a YAML file directly:

```{r}
expand_tines(schema, tmp_file)
```

# Case 3: Checking multiple alternatives of multiple steps

With `branch: multi` and more than one node, `expand_tines()` crosses every
node's choices together. 

```{r}
alts_factorial <- new_alternatives(
  branch = "multi", 
  node(
    overrides = "step-scaling",
    alternative(
      id = "step-scaling-zscore",
      decision = "z-score each variable",
      rationale = "puts variables on a common, interpretable scale"
    )
  ),
  node(
    overrides = "step-combine",
    alternative(
      id = "step-arithmetic-mean",
      decision = "use an arithmetic mean",
      rationale = "the old method"
    ),
    alternative(
      id = "step-weighted-mean",
      decision = "use a weighted mean, weighting each dimension by its variance",
      rationale = "down-weights noisier dimensions"
    )
  )
)
mv <- expand_tines(schema, alts_factorial)
mv
length(mv)
names(mv)
```

`step-scaling` has 1 alternative and `step-combine` has 2, so this crosses
into `(1 + 1) x (2 + 1)` = 6 schemas: `original`, `step-scaling-zscore`
alone, each of `step-combine`'s two alternatives alone, and each of those
two paired with `step-scaling-zscore`. 

