---
title: "Case Study: A Published GAITD Regression Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Case Study: A Published GAITD Regression Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE
)
```

```{=html}
<style>
pre, code {
  white-space: pre-wrap !important;
  overflow-wrap: break-word !important;
  word-break: break-word !important;
}
</style>
```

```{r setup}
library(tines)
```

This vignette walks through a real, published analysis end to end: Yee,
Frigau, and Ma (2025, *The Annals of Applied Statistics*) fit a GAITD
regression to model heaped smoking-initiation-age data. We use their first
response variable (smoking initiation age, "P1" in their paper) to show how
a schema can accompany a manuscript as reproducibility material, and how an
alternative file can encode a reviewer's suggested change.

# The schema

Yee et al.'s manuscript and supplementary material already state each
analytical decision and its rationale -- assembling a schema from it required
no new analysis, only collecting those decisions into the `tines` schema
format. We drafted a template with `draft_tines()` and filled it in by hand.

```{r results = "asis", echo = FALSE}
schema_path <- system.file("gaitd/schema.yml", package = "tines")
cat("<details>\n<summary>Click to read the filled schema.</summary>\n\n")
cat("```yaml\n")
cat(readLines(schema_path), sep = "\n")
cat("\n```\n\n</details>\n")
```

Consider the `special_value_mid_risk` step: the decision is to treat ages
12-14 and 19-21 as one inflated group with extra probability mass, since
these ages show a moderate, secondary peak in the data alongside the more
prominent peak at ages 15-18. Recording the rationale alongside the decision
is what lets a reader agree or disagree with the choice, and explore it, as
we do later in this vignette.

# Generating and validating the code

We read this into a `schema` object with `read_tines()`:

```{r}
schema <- read_tines(schema_path)
```

We then used `gen_code(schema, ...)` to translate this schema into an R script, and
`validate_script()` to fix any errors in the code generation. 

```{r results = "asis", echo = FALSE}
script_path <- system.file("gaitd/script.R", package = "tines")
cat("<details>\n<summary>Click to read the generated R script.</summary>\n\n")
cat("```r\n")
cat(readLines(script_path), sep = "\n")
cat("\n```\n\n</details>\n")
```

We can confirm the faithfulness of the schema and the generated script by running
the script and checking that it reproduces the estimation reported in the paper (Table 2 in Yee et al., 2025). The script produces a fitted model object, which we can summarize and display as a table:

```{r echo = FALSE}
source(script_path)
coef_table <- summary(fitted_model)@coef3
coef_df <- data.frame(
  term = rownames(coef_table),
  Estimate = coef_table[, "Estimate"],
  `Std. Error` = coef_table[, "Std. Error"],
  `z value` = coef_table[, "z value"],
  `Pr(>|z|)` = format.pval(coef_table[, "Pr(>|z|)"], digits = 3, eps = 0.001),
  row.names = NULL,
  check.names = FALSE
)
knitr::kable(coef_df, digits = 3)
```

# Exploring a reviewer's alternative

Yee et al.'s supplementary material also reports an investigation raised by a
reviewer: instead of pooling ages 12-14 and 19-21 into one inflated group,
treat 19-21 as a separate "alteration" group instead, since that older
subgroup shows a qualitatively different, declining pattern of uptake -- and
correspondingly drop to an intercept-only model, matching how the authors
computed AIC for this comparison. That is one coordinated change spanning
three steps, so it belongs in a `branch: single` alternative file. Again, you 
can draft a template with `draft_alternatives()` and fill it in by hand.

```{r results = "asis", echo = FALSE}
alt_path <- system.file("gaitd/alt-ap1921.yml", package = "tines")
cat("<details>\n<summary>Click to view the alternative YAML file.</summary>\n\n")
cat("```yaml\n")
cat(readLines(alt_path), sep = "\n")
cat("\n```\n\n</details>\n")
```

```{r}
alts <- read_alternatives(alt_path)
```

`expand_tines()` combines the schema and the alternative file into a
two-branch multiverse: the original analysis, and the reviewer's suggested
refit.

```{r}
mv <- expand_tines(schema, alts)
mv
```

We then used `gen_code()` again on the refit branch to generate an updated
script (again requiring an LLM, so not re-run here), and `validate_script()`
confirmed it reproduces the AIC of 19,340.4 that Yee et al. report for this
comparison. 

A reader who questions a choice does not need to wait for the authors to 
report that comparison: they can describe their alternative in natural
language in an alternative file, and let an LLM generate and validate the
corresponding code, to see directly how the result changes.

# References

Yee, T. W., Frigau, L., & Ma, C. (2025). Heaping and seeping, GAITD
regression and doubly constrained reduced-rank vector generalized linear
models in smoking studies. *The Annals of Applied Statistics*, 19(4),
3045-3070.

