---
title: "Structuring Schema Files"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Structuring Schema Files}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(tines)
```

A **schema** records a data analysis as an ordered sequence of decisions,
each written in natural language. It is the central object in `tines`: code
is generated from it, and it is what an alternatives file modifies to
produce a multiverse. This vignette covers

  * the anatomy of a schema YAML file,
  * the five ways to create one -- from a template, in R, from a file, from
    a data frame, or from prose with an LLM,
  * how a schema is mapped to a dataset, and
  * how pairing one with an alternatives file turns it into a multiverse.

# The big picture

A schema file has two parts: a `meta` block marking the document as a
schema, and a `nodes` list holding the steps of the analysis. The skeleton
always looks like this:

```yaml
meta:
  type: schema
  name: <optional label for the analysis>
nodes:
  - id: <step id>
    objective: <the problem this step solves>
    decision: <what was done>
    rationale: <why it was done that way>
  - id: <step id>
    ...
```

Each step carries four fields:

* **`id`** names the step. Ids must be unique within a schema, and they are
  how an alternatives file points at the step it overrides.
* **`objective`** states the problem the step addresses -- the question, not
  the answer.
* **`decision`** records what was actually done.
* **`rationale`** records why it was done that way. This is the field that
  lets a reader agree or disagree with a choice, rather than only observe it.

You can initialize the template above via `draft_tines()`, and 
fill in the steps by hand. Apart from that, there are four ways to create a schema: 


| If you are starting from | Use |
|----|----|
| nothing, and you would rather write YAML | `draft_tines()`, fill in the template, then `read_tines()` |
| nothing, and you would rather work in R | `build_schema()`, then one `add_step()` per decision |
| a schema file written earlier | `read_tines(path)` |
| a table of decisions, say from a spreadsheet | `as_schema(df)` |
| prose, such as a manuscript's methods section | `extract_schema(text, data_dict)` |

All five converge on the same thing: a `schema` object, which is a tibble
of steps. The YAML file is the format you store and share; the object is
what you work with in R. Most functions that consume a schema (`expand_tines()`, `gen_code()`,
`draw_tines()`) take either. Below we show examples of creating a schema in each of the five ways.

# Creating a schema

## From a template

`draft_tines()` writes a skeleton with two placeholder steps, ready to fill
in by hand:

```{r}
draft_path <- withr::local_tempfile(fileext = ".yml")
draft_tines(file_path = draft_path)
```

```{r results = "asis", echo = FALSE}
cat("```yaml\n")
cat(readLines(draft_path), sep = "\n")
cat("\n```\n")
```

## In R

`build_schema()` starts an empty schema and `add_step()` appends to it, one
step per decision:

```{r}
hdi <- build_schema(name = "HDI Example") |>
  add_step(
    id = "step-scaling",
    objective = "variables are in different scales",
    decision = "apply min-max scaling to each variable",
    rationale = "to put them on the same scale for combination"
  ) |>
  add_step(
    id = "step-combine",
    objective = "combine the three dimensions into a single index",
    decision = "use the geometric mean",
    rationale = "the geometric mean penalizes uneven development"
  )
hdi
```

`write_tines()` writes it back out as YAML:

```{r}
out_path <- withr::local_tempfile(fileext = ".yml")
write_tines(hdi, out_path)
```

```{r results = "asis", echo = FALSE}
cat("```yaml\n")
cat(readLines(out_path), sep = "\n")
cat("\n```\n")
```

## From a file

Most often the schema already exists. `read_tines()` reads one in -- here
the Human Development Index example shipped with the package, which
combines health, education, and standard of living into a single index:

```{r results = "asis", echo = FALSE}
schema_path <- system.file("hdi.yml", package = "tines")
cat("```yaml\n")
cat(readLines(schema_path), sep = "\n")
cat("\n```\n")
```

```{r}
schema <- read_tines(schema_path)
schema
```

Because the result is a tibble, the decisions of an analysis can be
inspected with the tools you already use for data frames:

```{r}
schema$id
schema$rationale[schema$id == "step-combine"]
```


## From a data frame

Decisions are often catalogued somewhere else first -- a spreadsheet, or a
table in a lab notebook. Any data frame carrying the four step fields
coerces directly:

```{r}
steps <- data.frame(
  id = c("step-clean", "step-model"),
  objective = c("handle missing values", "estimate the effect"),
  decision = c("drop incomplete cases", "fit a linear model"),
  rationale = c("listwise deletion keeps the sample interpretable",
                "the relationship is assumed linear")
)

as_schema(steps, name = "From a spreadsheet")
```

`inputs` and `outputs` may be supplied as list-columns if you have them;
otherwise add them later with `update_io()`. A data frame missing any of
the four required fields is refused, and says which:

```{r error = TRUE}
as_schema(steps[, c("id", "decision")])
```

## From a manuscript, with an LLM

`extract_schema()` drafts a schema from a prose description of the
methodology, such as the methods section of a paper:

```{r eval = FALSE}
# Requires an LLM API key (e.g. ANTHROPIC_API_KEY); not run automatically.
extract_schema(
  text = football_grp5,
  data_dict = c("player", "redCards", "rater1", "rater2"),
  output_file = "draft_schema.yml"
)
```

A schema drafted this way carries extra bookkeeping fields -- `confidence`
and `clarification_question` -- recording how sure the model was that it read
each decision correctly. Any extra field is preserved as
its own column when the file is read, and written back out unchanged:

```{r}
drafted <- example_football_grp5()
colnames(drafted)
```

# Mapping a schema to data

The four core fields describe an analysis, but generating code that runs
against a real dataset also requires knowing which variables each step
consumes and produces. Two optional fields record that:

```yaml
  - id: step-scaling
    objective: variables are in different scales
    decision: apply min-max scaling to each variable
    rationale: to put them on the same scale for combination
    inputs: [life_exp, exp_sch, avg_sch, gni]
    outputs: [scaled]
```

* **`inputs`** name the variables the step reads.
* **`outputs`** name the variables it creates. An output need not exist in
  the dataset yet -- the schema describes the analysis before its code
  exists.

Supply a dataset and `tines` checks the mapping. It walks the steps in
order, starting from the dataset's own columns, and requires every step's
`inputs` to be available by that point -- either a column of the original
data, or an output created by an earlier step:

```{r}
hdi_data <- data.frame(
  life_exp = c(72, 81, 66),
  exp_sch = c(13, 16, 10),
  gni = c(9000, 42000, 3200)
)

mapped <- build_schema(name = "HDI", data = hdi_data) |>
  add_step(
    id = "step-scaling",
    objective = "variables are in different scales",
    decision = "apply min-max scaling to each variable",
    inputs = c("life_exp", "exp_sch", "gni"),
    outputs = "scaled"
  ) |>
  add_step(
    id = "step-combine",
    objective = "combine the dimensions into a single index",
    decision = "use the geometric mean",
    inputs = "scaled",
    outputs = "hdi"
  )
```

`step-combine` reads `scaled`, which is not a column of `hdi_data` but is
produced by `step-scaling`, so the mapping is valid.

Swapping in a dataset that breaks the mapping is refused outright, naming
the step and the variables it can no longer reach:

```{r error = TRUE}
renamed <- data.frame(
  life_exp = c(70, 80, 60),
  exp_sch = c(12, 15, 9),
  income = c(8000, 40000, 3000) # `gni` renamed
)
update_data(mapped, renamed)
```

Three functions maintain the mapping. `update_data()` attaches a different
dataset and re-validates the whole schema against it, as above.
`update_io()` edits the mapping for a single step; it warns rather than
refuses when a variable is unreachable, so you can fix several steps in
sequence without the intermediate states blocking you. `gen_io()` fills in
the mapping for every step at once by asking an LLM to match the steps
against the columns of a dataset.

# From one schema to many

A schema describes one analysis. Pairing it with an **alternatives file** --
which records different ways of carrying out one or more of its steps --
and passing both to `expand_tines()` produces a `multiverse`, one schema per
branch the alternatives file describes, alongside the original:

```{r}
alts <- read_alternatives(system.file("hdi-alt-single.yml", package = "tines"))
expand_tines(schema, alts)
```

See `vignette("alternatives")` for the anatomy of an alternatives file and
the two ways it can be expanded.
